Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism

### 2024-07-23 _tex/index.tex
Built site for gh-pages

In contemporary contexts, Thurstone's law of comparative judgment
\citeyearpar{Thurstone_1927} primarily refers to the method of
\emph{Dichotomous} Comparative Judgment
\citep[DCJ,][]{Pollitt_2012a, Pollitt_2012b}. In DCJ, a judge assesses
the relative manifestation of a \emph{trait} within a pair of stimuli.
This assessment results in a dichotomous value indicating which stimulus
possesses a higher degree of the trait. After different judges perform
multiple rounds of pairwise comparisons, an outcome vector is produced.
This vector is modeled using the Bradley-Terry-Luce model
\citep[BTL,][]{Bradley_et_al_1952, Luce_1959}, which creates a score
that corresponds with the trait of interest. This score is then used to
rank the stimuli from lowest to highest or to evaluate the influence of
certain variables on the stimuli's positions in the ranking.

DCJ has proven effective in assessing competencies and traits
predominantly within the educational realm, as demonstrated by
\citet{Pollitt_2012b}, \citet{Jones_2015}, \citet{vanDaal_et_al_2016},
\citet{Bartholomew_et_al_2018}, \citet{Lesterhuis_2018},
\citet{Bartholomew_et_al_2020}, and \citet{Marshall_et_al_2020}.
However, its application transcends education, as exemplified by
\citet{Boonen_et_al_2020}. The methodology has also evolved to include
multiple, as opposed to pairwise comparisons
\citep{Luce_1959, Placket_1975}, and to accommodate comparisons with
ordinal outcomes \citep{Tutz_1986, Agresti_1992}. Overall, research
suggests that DCJ offers an alternative and efficient approach to
measurement and evaluation, characterized by its reliability and
validity \citep{Lesterhuis_2018, vanDaal_2020, Marshall_et_al_2020}.
Nevertheless, despite the method's widespread use, the literature does
not offer a clear representation of the plausible mechanisms that
generate the DCJ data. Particularly, there is no depiction of the
complexity and the underlying assumptions of the DCJ system, nor how
different assessment factors can potentially influence the observed DCJ
outcome.

According to \citet{Verhavert_et_al_2019} and \citet{vanDaal_2020},
several assessment factors interact and influence the DCJ outcome. These
factors include the number and characteristics of the stimuli, their
\emph{proximity} in terms of the assessed trait, the number of
comparison per stimulus, and the pairing algorithm used. Furthermore,
since the method relies on judges' assessments, the number and
characteristics of judges, their \emph{discrimination} abilities, and
the number of comparisons per judge also play pivotal roles. Moreover,
when the stimuli represent sub-units of higher-levels units, factors
such as the number and characteristics of these units, along with their
\emph{proximity} in terms of the assessed trait, can significantly
influence the outcome. For instance, \citet{vanDaal_et_al_2016} assessed
university students' skills in academic writing, utilizing multiple
argumentative essays (stimuli, sub-units) originating from various
students (units).

Although several studies have examined the individual impact of these
factors on the method's reliability
\citep{Bramley_2015, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, vanDaal_et_al_2017, Gijsen_et_al_2021},
none, to the best of the authors' knowledge, have provided a transparent
depiction of the DCJ system and the plausible mechanisms generating the
DCJ outcome. This study aims to fill this gap by representing DCJ within
the causal analysis framework \citep{Pearl_2009, Pearl_et_al_2016}.
Using Directed Acyclic Graphs
\citep[DAG,][]{Gross_et_al_2018, Neal_2020}, the study will construct a
scientific model to elucidate the underlying assumptions of the DCJ
system, providing plausible mechanisms for how the DCJ outcome could be
generated. Next, using a minimal set of assumptions embedded in the
framework, the study will translate the scientific model into a
probabilistic statistical model. This model derives statistical
estimands to draw inferences about plausible causal relationships within
the DCJ system. Ultimately, this research aims to extend the law of
comparative judgment initially proposed by \citet{Thurstone_1927} and
provide a sound probabilistic base for the statistical analysis of DCJ
data. Consequently, this research holds significance for researchers and
analysts involved in education and assessment procedures who implement
or design DCJ experiments.

@@

\subsection{Causal analysis and DAGs}\label{sec-prelim-causal}

Causal analysis is the field of experimental design and statistics
pertaining to establishing cause and effect from data.

Causal inference

Directed Acyclic Graph (DAG)

the four components of DAGs

Structural Causal Model (SCM)

the four components of DAGs in SCM notation

@@

\subsection{Where does it all fit?}\label{sec-theory-where}

@@

\subsection{Why do we need to estimate judges'
abilities?}\label{sec-appA}

>>>>

In contemporary contexts, Thurstone's law of comparative judgment
\citeyearpar{Thurstone_1927} primarily refers to the method of
\emph{dichotomous} comparative judgment
\citep[DCJ,][]{Pollitt_2012a, Pollitt_2012b}. In DCJ, a judge assesses
the relative manifestation of a \emph{trait} within a pair of stimuli.
This assessment results in a dichotomous value indicating which stimulus
possesses a higher degree of the trait. After different judges perform
multiple rounds of pairwise comparisons, an outcome vector is produced.
This vector is modeled using the Bradley-Terry-Luce model
\citep[BTL,][]{Bradley_et_al_1952, Luce_1959}, which creates a score
that corresponds with the trait of interest. This score is then used to
rank the stimuli from lowest to highest or to evaluate the influence of
certain variables on the stimuli's positions in the ranking.

DCJ has proven effective in assessing competencies and traits
predominantly within the educational realm, as demonstrated by
\citet{Pollitt_2012b}, \citet{Jones_2015}, \citet{vanDaal_et_al_2016},
\citet{Bartholomew_et_al_2018}, \citet{Lesterhuis_2018},
\citet{Bartholomew_et_al_2020}, and \citet{Marshall_et_al_2020}.
However, its application transcends education, as exemplified by
\citet{Boonen_et_al_2020}. The methodology has also evolved to include
multiple, as opposed to pairwise comparisons
\citep{Luce_1959, Placket_1975}, and to accommodate comparisons with
ordinal outcomes \citep{Tutz_1986, Agresti_1992}. Overall, research
suggests that DCJ offers an alternative and efficient approach to
measurement and evaluation, characterized by its reliability and
validity \citep{Lesterhuis_2018, vanDaal_2020, Marshall_et_al_2020}.
Nevertheless, despite the method's widespread use, the literature does
not offer a clear representation of the plausible mechanisms that
generate DCJ data. Particularly, there is no depiction of the complexity
and the underlying assumptions of the DCJ system, nor how different
assessment factors can potentially influence the observed DCJ outcome.

According to \citet{Verhavert_et_al_2019} and \citet{vanDaal_2020},
several assessment factors interact and influence the method's outcome.
These factors include the number and characteristics of the stimuli,
their \emph{proximity} in terms of the assessed trait, the number of
comparison per stimulus, and the pairing algorithm used. Furthermore,
since the method relies on judges' assessments, the number and
characteristics of judges, their \emph{discrimination} abilities, and
the number of comparisons per judge also play pivotal roles. Moreover,
when the stimuli represent sub-units of higher-levels units, factors
such as the number and characteristics of these units, along with their
\emph{proximity} in terms of the assessed trait, can significantly
influence the outcome. For example, \citet{vanDaal_et_al_2016} assessed
the academic writing skills of university students (units) using
multiple argumentative essays (sub-units).

Although several studies have examined the individual impact of these
factors on the method's reliability
\citep{Bramley_2015, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, vanDaal_et_al_2017, Gijsen_et_al_2021},
none, to the best of the authors' knowledge, have provided a transparent
depiction of the DCJ system and the mechanisms generating the DCJ
outcome. This study aims to fill this gap by representing DCJ within the
causal inference framework. Specifically, using the \emph{structural
approach} to causal inference
\citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}, the study aims to
construct a scientific model. This model will elucidate the underlying
assumptions of the DCJ system, providing plausible mechanisms for how
the DCJ outcome is generated. Next, using a minimal set of assumptions,
the study will translate the scientific model into a probabilistic
statistical model. This model will produce statistical estimates to draw
inferences about plausible causal relationships within the DCJ system.

Ultimately, this research aims to extend the law of comparative judgment
initially proposed by \citet{Thurstone_1927} and provide a sound
probabilistic base for the statistical analysis of DCJ data.
Consequently, this research holds significance for researchers and
analysts involved in education and assessment procedures who implement
or design DCJ experiments.

@@

\subsection{The structural approach to causal
inference}\label{sec-prelim-structural}

In statistics, \emph{causal inference} refers to the process of
identifying the causes of a phenomenon and estimating their effects
using data \citep{Shaughnessy_et_al_2010, Neal_2020}. Unlike classical
statistical modeling, which focuses solely on summarizing data and
inferring associations, causal inference provides a coherent
mathematical representation for analyzing causes and counterfactuals
\citep{Pearl_2009}. Counterfactuals reflect scenarios \emph{contrary to
fact}, where alternative \emph{potential} outcomes have not been
observed and cannot be observed \citep{Neal_2020, Counterfactual_2024}.
Using counterfactuals, researchers can construct a theory of the world
that explains why specific causes have specific effects, and what
happens in the absence of those causes. Consequently, counterfactuals
represent the highest level of abstraction in causal inference
\citep{Pearl_et_al_2018}.

Several approaches to constructing counterfactuals and applying causal
inference exist, but two are particularly prominent: the potential
outcomes approach and the structural approach. The potential outcomes
approach, also known as the Neyman-Rubin causal model, was developed by
\citet{Neyman_et_al_1990} and \citet{Rubin_1974}. In contrast, the
structural approach was pioneered by \citet{Wright_1921}, and later
extended by Pearl \citep{Pearl_2009, Pearl_et_al_2016}. Both approaches
employ rigorous mathematical notation to characterize causal inference,
but they do so in different ways \citep{Neal_2020}. The potential
outcomes approach relies on counterfactual notation, whereas the
structural approach utilizes the do-operator and Structural Causal
Models \citep[SCM,][]{Pearl_2009, Pearl_et_al_2016}. Despite these
differences, both approaches provide methods for using experimental and
observational data to estimate causal effects \citep{Pearl_2010}.

However, the structural approach offers a significant advantage over the
potential outcomes approach by enabling the graphical representation of
systems through directed acyclic graphs
\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. These graphical
representations provides a transparent depiction of a system's
complexity, revealing its underlying assumptions and the plausible
mechanisms generating the system's outcome. This capability aligns well
with the aims of this study. Consequently, the structural approach will
be the primary method used to address the study's research goal.

\subsection{Graphs, DAGs and SCMs}\label{sec-prelim-graphs}

@@

@@

\subsection{Additional definitions}\label{sec-appA}

\textbf{Counterfactual:}

\subsection{Why do we need to estimate judges'
abilities?}\label{sec-appB}
